Evolutionary algorithms in a volunteer computing system: performance assessment

### 2016-03-11 PPSN2016/paper.Rnw
Completely rewritten introduction
Refs #109

Please @mariosky @unintendedbear @fergunet take a look; it describes the approach and what we are going to do in the paper and introduces new references. It probably should be shorter, because there is a 10 page limit, divided thus

* 1 page: intro
* 1 page: SoA
* 2 pages: Description framework
* 3 pages: results
* 1 page: conclusion
* 2 pages: Acknowledgements and references.

the 10 page limit is hard to do...

Some time ago, our group faced the problem of diminishing funds for
buying new hardware. This was aggravated by the increasing maintenance
costs and extended downtime resulting from the continuous failures of
existing clusters.  Considering this, we leveraged our experience in
the design of web applications with JavaScript and other volunteer and
unconventional distributed evolutionary computing systems to design
and release a new free framework that would allow anyone to create a
volunteer distributed evolutionary computation (EC) experiment using
cloud resources as servers and browsers as clients. This framework was
called NodIO \cite{2016arXiv160101607M}. NodIO provides server
infrastructure for volunteer-based distributed evolutionary computing
experiments by providing a chromosome {\em pool}. This pool is used by
clients in browsers and any other using the application programming
interface (API) to put chromosomes and retrieve them, working then as
a loose, asynchronous and {\em ad hoc} connection among all clients
using it.

This loose connection provides a low-overhead way to connect desktop
experiments with volunteer-based ones, with all of them contributing
to the pool, but every one of them working as separate island carrying
out their own evolutionary algorithms. This is why NodIO is proposed
mainly as a {\em complement} to existing resources such as desktop
systems or laptops. As long as it provides a non-null computational
capability that can help existing resources find the solution faster
it will have found its purpose. The main use case is someone setting
up NodIO in the cloud, writing a fitness evaluation function and
running a client from his or her own computer, but requesting help in
social networks for additional resources. That is why, in this system,
we have considered the whole {\em social} aspect in the design, with
issues related to security, trust and privacy among others. The
computing system becomes a {\em sociotechnical} system
\cite{vespignani2009predicting}. In this paper, we are going to focus
on measuring the response of users to experiments, that is, the time
they spend running it, but at the same time we will also focus on the
technical aspects of the server and how these might change the
behavior of users, improving the capability of the system.

Our research group is committed to open science, and we think this is
a very important part of the techno-social system. By being
transparent, incentives to cheat are reduced and, in fact, we have
detected no issue for the time being. Next we present the state of the
art in web-based volunteer computing systems along with attempts to
predict and model its behavior.

>>>>

The world wide web provides not only a platform for content
distribution, but an increasingly reliable and high-performance
platform for running distributed applications. This is based in two
key facts: the Javascript virtual machine every browser runs and a
simplified standard interface for interacting with servers. Thus
creating a distributed computing experiment is just a matter of making
a JavaScript application interchange information with a server, called
rest. From the point of view of the programmer, this involves
relatively common skills and no special libraries, since that
interface is built in the browser, and a simple application that
responds to those requests on the server side; both involve just a few
dozens lines of code additionally to whatever business logic your
application has. But, more importantly and from the point of view of
the user, that application can be run by simply visiting a web page.

Using this system for creating distributed experiments is called {\em
  volunteer}, {\em cycle-scavenging} or {\em opportunistic} computing
\cite{sarmenta2001volunteer} and it dates back, in different shapes
and underlying mechanisms, to the origin of the web \cite{david-seti:home}. Our interest
here, however, is to use it as a resource for evolutionary
computation. This has been a line of research in our group for a long
time \cite{jj-ppsn98}, but has lately been revisited thanks to the
ubiquity of social networks and handheld computing devices.

In this line of research that uses volunteer computing for
evolutionary algorithms, there are several pending issues. The first
and maybe most important is approaching volunteer computing as a
techno-social system \cite{vespignani2009predicting} which integrates
user decisions and behavioral patterns in the system model; this
includes trying to optimize the number of users in a particular
experiment. The second line of research, although related to the
first, is more focused on the evolutionary algorithm itself and how
different design decisions will affect its performance.

In this paper our focus will be in the second aspect: we will try to
design a decentralized system that, at the same time, is able to use
all resources available. This design will be done incrementally by
changing the system on the client and the server and measuring its
impact on the overall performance: time and evaluations needed to find
the solution. Eventually, we want to find a system that, whatever the
number of users available to perform the experiment, is able to
maximize their contribution to the evolutionary algorithm, at the same
time that the evolutionary algorithm itself makes the most of those
contributions, independently of how many they are.

The rest of the paper is organized as follows: Next we will briefly
present the state of the art in opportunistic distributed evolutionary
computation (EC). Section \ref{sec:description} will describe the
framework and problem used in the experiments, which are publicly
available under a free license. We will present the results of the
different steps in the incremental design in Section
\ref{sec:experiments}, to finally wrap up with the conclusions.
